---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the repdata-data-StormData.csv.bz2 file into workspace.
```{r}

library("R.utils")
if (!file.exists("./repdata-data-StormData.csv")) {        
        fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        dest <- "./repdata-data-StormData.csv.bz2"
        download.file(fileURL, dest)
        bunzip2("./repdata-data-StormData.csv.bz2")
}
```
        
## Read data
Read the repdata-data-Stormdata.csv into data frame.
```{r}
data <- read.table("./repdata-data-StormData.csv", header = TRUE, sep = ",", na.strings = "", comment.char = "#")
cnames <-readLines("./repdata-data-StormData.csv", 1)
cnames <- strsplit(cnames, ",", fixed = TRUE)
cnames <- make.names(cnames[[1]])
cnames<-gsub("(.*)\\.", "\\1", cnames)
cnames<-gsub("_", "", cnames)
names(data) <- cnames
```
## Which types of events are most harmful with respect to population health?
Convert data to data table.
```{r}
library("data.table")
dt <- data.table(data)
```
Summarise fatalities then group by event type. Order by sum and display the top five items.
```{r}
dataFatal <- dt[, list(sum=sum(X.FATALITIES)), by = X.EVTYPE]
dataFatal <- dataFatal[rev(order(dataFatal$sum))]
head(dataFatal)
dataFatal10 <- dataFatal[1:10]
```
We can see that tornado is causing the highest sum of fatalities with 5633 fatalities.

Summarise injuries then group by event type. Order by sum and display the top five items.
```{r}
dataInjury <- dt[, list(sum=sum(X.INJURIES)), by = X.EVTYPE]
dataInjury <- dataInjury[rev(order(dataInjury$sum))]
head(dataInjury)
dataInjury10 <- dataInjury[1:10]
```
We can see that tornado is causing the highest sum of injuries with 91346 injuries.

Summarise both fatalities and injuries and group by event type. Order by sum and display the top five items.
```{r}
dataCombined <- dt[, list(sum=sum(X.FATALITIES, X.INJURIES)), by = X.EVTYPE]
dataCombined <- dataCombined[rev(order(dataCombined$sum))]
head(dataCombined)
dataCombined10 <- dataCombined[1:10]
```
We can see that Tornado is causing the highest total sum of fatalities and injuries with 96979 occurences.

## Which types of events have the greatest economic consequences?
Let us load the plyr package.
```{r}
library(plyr)
```

Let us extract dataProp and dataCrop containing the data needed for analysis.
```{r}
dataProp <- ddply(data, .(X.EVTYPE,X.PROPDMGEXP),function(x) sum(x$X.PROPDMG))
colnames(dataProp)[3] <- "Count"
dataCrop <- ddply(data, .(X.EVTYPE,X.CROPDMGEXP),function(x) sum(x$X.CROPDMG))
colnames(dataCrop)[3] <- "Count"
```

Let us summarize dataProp and dataCrop.
```{r}
summary(dataProp)
```

```{r}
summary(dataCrop)
```

Let us do some data cleaning by removing null and empty values.
```{r}
dataProp <- dataProp[complete.cases(dataProp$X.PROPDMGEXP),]
dataCrop <- dataCrop[complete.cases(dataCrop$X.CROPDMGEXP),]
```

Substitute exponential values (H/K/M/B)
```{r}
dataProp$X.PROPDMGEXP<-sapply(as.character(dataProp$X.PROPDMGEXP),switch,'M'=10^6,'m'=10^6,'K'=10^3,'k'= 10^3,'B'=10^9,'0' = 0, '2' = 2)
dataCrop$X.CROPDMGEXP<-sapply(as.character(dataCrop$X.CROPDMGEXP),switch,'M'=10^6,'m'=10^6,'K'=10^3,'k'= 10^3,'B'=10^9,'h'= 100, 'H' = 100, '0'=0,'1'=1,'2'=2,'3'=3,'4'=4,'5'=5,'6'=6,'7'=7, '8'=8)
```

Remove NULL values
```{r}
dataProp <- dataProp[!sapply(dataProp$X.PROPDMGEXP,is.null),]
dataCrop <- dataCrop[!sapply(dataCrop$X.CROPDMGEXP,is.null),]
```

Calculate total property damage for crop and prop.
```{r}
totalPropDamage <- ddply(dataProp, .(X.EVTYPE), function(x) sum(as.numeric(x$X.PROPDMG)*x$Count))
totalCropDamage <- ddply(dataCrop, .(X.EVTYPE), function(x) sum(as.numeric(x$X.CROPDMG)*x$Count))
colnames(totalPropDamage)[2] <- "Count"
colnames(totalCropDamage)[2] <- "Count"
```

Calculate the combined damage
```{r}
totalCombinedDamage <- rbind(totalCropDamage,totalPropDamage)
totalCombinedDamage <- ddply(totalCombinedDamage,.(X.EVTYPE), function(x) sum(x$Count))
colnames(totalCombinedDamage)[2] <- "Count"
```

Sort our data
```{r}
sortedPropDamage <- totalPropDamage[with(totalPropDamage,order(Count,decreasing = TRUE)),]
sortedCropDamage <- totalCropDamage[with(totalCropDamage,order(Count,decreasing = TRUE)),]
sortedCombinedDamage <- totalCombinedDamage[with(totalCombinedDamage,order(Count,decreasing = TRUE)),]
```

Let us have a look at top five for our datasets
```{r}
sortedPropDamage10 <- sortedPropDamage[1:10,]
head(sortedPropDamage10)
```
We can see that for prop flood caused 145 Billion USD damage.

```{r}
sortedCropDamage10 <- sortedCropDamage[1:10,]
head(sortedCropDamage10)
```
We can see that for crop drought caused 14 Billion USD damage.

```{r}
sortedCombinedDamage10 <- sortedCombinedDamage[1:10,]
head(sortedCombinedDamage10)
```
We can see that for the combined damage (crop + prop) flood caused 150 Billion USD damage.

## Results

### Population health
From our analysis we can conclude that the most dangerous weather event was tornado with 5633 fatalities and 91345 injuries.
```{r}
library(ggplot2)
library(grid)
install.packages("gridExtra", repos="http://cran.rstudio.com/")
library(gridExtra)

plotTop10Health <- ggplot(dataCombined10,aes(x= X.EVTYPE,y = sum))
plotTop10Health <- plotTop10Health+geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Weather Event") + ylab("Sum of Fatalities and injuries") 
grid.arrange(plotTop10Health,ncol = 1)
```

### Population health
From our analysis we can conclude that flood caused the highest combined damage with 150 Billion USD.
```{r}
plotTop10Damage <- ggplot(sortedCombinedDamage10,aes(x= X.EVTYPE,y = Count))
plotTop10Damage <- plotTop10Damage+geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Weather event") + ylab("Sum of combined crop and prop damage") 
grid.arrange(plotTop10Damage,ncol = 1)
```